---
title: "Geographical Segmentation with Spatially Constrained Clustering Techniques"
editor: visual
execute: 
  error: false
  warning: false
---

## Overview

Grouping areas based on their proximity and attributes is necessary for business decisions. By doing so, we can derive areas with similar attributes to focus or apply different strategies. For some examples, refer to [this article](https://manychat.com/blog/geographic-segmentation/).

In this exercise, we will practice how to delineate homogeneous region by using geographically referenced multivariate data. There are two major analysis, namely:

-   hierarchical cluster analysis; and

-   spatially constrained cluster analysis.

## Getting Started

### The analytical question

In this hands-on exercise, we are interested to delineate [Shan State](https://en.wikipedia.org/wiki/Shan_State), [Myanmar](https://en.wikipedia.org/wiki/Myanmar) into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

### The data

Two data sets will be used in this exercise:

-   **Myanmar Township Boundaries**: a ESRI shapefile consisting of Myanmar townships' boundary. The data is captured in polygon feature.

-   **Shan-ICT.csv**: this is the data captured from the [2014 Population and Housing Census of Myanmar](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet).

### Installing and loading R packages

The R packages needed for this exercise are as follows:

-   Spatial data handling

    -   **sf**, **rgdal** and **spdep**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

-   Multivariate data visualisation and analysis

    -   **corrplot**, **ggpubr**, and **heatmaply**

-   Cluster analysis

    -   **cluster**

    -   **ClustGeo**

```{r}
pacman::p_load(sf, rgdal, spdep, tidyverse,
               tmap, corrplot, ggpubr, heatmaply,
               factoextra, cluster, ClustGeo, NbClust, psych)
```

## Data Import and Preparation

### Importing geospatial data into R environment

We will use `st_read()` to import the shapefile of Myanmar township's boundary. Using `filter()`, we will only import the data with `ST` values of Shan (East), Shan (North), Shan (South). They are the areas belonging to the Shan state in Myanmar.

```{r}
shan_sf <- st_read(dsn = "data3/geospatial", layer = "myanmar_township_boundaries") %>% 
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

`shan_sf` is a **simple feature data.frame** file. Let's view the content of the file. Notice that sf.data.frame is conformed to Hardy Wickham's [tidy](https://edzer.github.io/rstudio_conf/#1) framework. Therefore, `glimpse()` can also be used to reveal the few first fields of the data frame.

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

### Importing aspatial data into R environment

```{r}
ict <- read.csv("data3/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R\'s \* tibble data.frame\* format.

Let's view the summary of `ict`.

```{r}
summary(ict)
```

Notice that there are 55 fields in this data set, matching the 55 areas in `shan_sf`.

### Deriving new variables using `dplyr` package

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the penetration rate of each ICT variable by using the code chunk below.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total.households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total.households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land.line.phone`/`Total.households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile.phone`/`Total.households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total.households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet.at.home`/`Total.households`*1000) %>%
  rename(`DT_PCODE` =`District.Pcode`,`DT`=`District.Name`,
         `TS_PCODE`=`Township.Pcode`, `TS`=`Township.Name`,
         `TT_HOUSEHOLDS`=`Total.households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land.line.phone`, `MPHONE`=`Mobile.phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet.at.home`) 
```

Let's review the summary of `ict_derived`.

```{r}
summary(ict_derived)
```

## EDA

### EDA using statistical graphics

We will use histogram and boxplot to examine the distribution and outliers of the variables.

```{r}
radiohist <- ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
radiobox <- ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
ggarrange(radiohist, radiobox, nrow = 2, ncol = 1)
```

Next, we will also plot the distribution of the radio percentage `RADIO_PR`.

```{r}
radio <- ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
radioprbox <- ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
ggarrange(radio, radioprbox, nrow = 2, ncol = 1)
```

The distribution of `RADIO` is skewed to the right with 3 outliers, while the distribution of `RADIO_PR` is less right-skewed and has less outlier.

Now, let's look at the distribution of other newly derived variables.

```{r}
tv <- ggplot(data=ict_derived, 
       aes(x=`TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
llphone <- ggplot(data=ict_derived, 
       aes(x=`LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
mphone <- ggplot(data=ict_derived, 
       aes(x=`MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
computer <- ggplot(data=ict_derived, 
       aes(x=`COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
internet <- ggplot(data=ict_derived, 
       aes(x=`INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
ggarrange(radio, tv, llphone, mphone, computer, internet,
          nrow = 2, ncol = 3)
```

### EDA using choropleth map

#### Joining geospatial data with aspatial data

To prepare the choropleth map, we need to combine both geospatial and aspatial data together. The function used for that task is `left_join()` of **dplyr** package. The *shan_sf* simple feature data.frame will be used as the base data object and the *ict_derived* data.frame will be used as the join table.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, 
                     by=c("TS_PCODE"="TS_PCODE"))
```

#### Preparing a choropleth map

Let's have a quick look at the distribution of radio penetration `RADIO_PR` of Shan State at township level.

```{r}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the total number of households (i.e. TT_HOUSEHOLDS.map) and one for the total number of household with Radio (RADIO.map) by using the code chunk below.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) +
  tm_layout(legend.width = 0.3,
            legend.height = 0.3,
            legend.text.size = 0.5)

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) +
  tm_layout(legend.width = 0.3,
            legend.height = 0.3,
            legend.text.size = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

We can see that where the total number of households is high, the number of households with radio is high too.

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate by using the code chunk below.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

With radio penetration `RADIO_PR`, the radio distribution is no longer correlate with total household distribution.

## Correlation Analysis

Before we perform cluster analysis, it is important for us to ensure that the cluster variables are not highly correlated.

In this section, you will learn how to use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise and analyse the correlation of the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

## Hierarchy Cluster Analysis

### Extrating clustering variables

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Notice that the final clustering variables list does not include variable INTERNET_PR because it is highly correlated with variable COMPUTER_PR.

Next, we need to change the rows by township name instead of row number by using the code chunk below.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Now we will delete the `TS.x` column by using select and assign the data table without `TS.x` to `shan_ict`.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

### Data Standardization

In general, multiple variables will be used in cluster analysis. It is not unusual their values range are different. In order to avoid the cluster analysis result is biased to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

#### Min-Max standardization

`normalize()` of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package is used to stadardisation the clustering variables by using Min-Max method. Notice that the values range of the Min-max standardised clustering variables are 0-1 now.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

#### Z-score standardization

Z-score standardization can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. Notice the mean and standard deviation of the Z-score standardized clustering variables are 0 and 1 respectively. [*describe()*](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**psych**](https://cran.r-project.org/web/packages/psych/) package is used here instead of *summary()* of Base R because the earlier provides standard deviation.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

***Warning: Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.***

In this case, the histograms of `LLPHONE_PR`, `MPHONE_PR` and `COMPUTER_PR` are right-skewed, so it is unlikely that the distribution of those variables is normal. For further analysis, we can also conduct distribution fit test to find out.

#### Visualising the standardized clustering variables

Beside reviewing the summary statistics of the standardized clustering variables, it is also a good practice to visualize their distribution graphical.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw Value")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

Notice that the overall distribution of the clustering variables will change after the data standardization. Hence, it is advisible **NOT** to perform data standardization if the values range of the clustering variables are not very large.

### Computing proximity matrix

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**. The default is *euclidean* proximity matrix.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
proxmat
```

### Computing hierarchical clustering

In R, there are several packages provide hierarchical clustering function. In this hands-on exercise, [*hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) of R stats will be used.

*hclust()* employed agglomeration method to compute the cluster. Eight clustering algorithms are supported, they are: ward.D, ward.D2, single, complete, average(UPGMA), mcquitty(WPGMA), median(WPGMC) and centroid(UPGMC).

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

Using `plot()` of Base R, we can plot the cluster dendogram.

```{r}
plot(hclust_ward, cex = 0.6)
```

### Selecting the optimal clustering algorithm

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

Ward\'s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward\'s method will be used.

### Determining Optimal Clusters

Another technical challenge face by data analyst in performing clustering analysis is to determine the optimal clusters to retain.

There are [three](https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/) commonly used methods to determine the optimal clusters, they are:

-   [Elbow Method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   [Average Silhouette Method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   [Gap Statistic Method](https://statweb.stanford.edu/~gwalther/gap.pdf)

#### Gap Statistic Method

The [**gap statistic**](http://www.web.stanford.edu/~hastie/Papers/gap.pdf) compares the total within intra-cluster variation for different values of k with their expected values under null reference distribution of the data. The estimate of the optimal clusters will be value that maximize the gap statistic (i.e., that yields the largest gap statistic). This means that the clustering structure is far away from the random uniform distribution of points.

To compute the gap statistic, [*clusGap()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package will be used. Also note that the [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

```{r}
fviz_gap_stat(gap_stat)
```

The option of 1 cluster has the biggest gap. However, having just 1 cluster does not make sense for the purpose of clustering. Among the other options, 6 cluster has the highest gap statistic. **We will choose 6 clusters.**

### Interpreting the dendrograms

The height of the fusion, provided on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually-driven hierarchical clustering analysis

In this section, we will learn how to perform visually-driven hiearchical clustering analysis by using [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package.

With **heatmaply**, we are able to build both highly interactive cluster heatmap or static cluster heatmap.

#### Transforming the data frame into a matrix

The data `shan_ict` is a data frame, so we need to transform it to a matrix to draw the heatmap.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting interactive cluster heatmap using *heatmaply()*

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Mapping the clusters formed

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;

-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and

-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Let's draw the map with cluster!

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

The choropleth map above reveals the clusters are very fragmented. The is one of the major limitation when non-spatial clustering algorithm such as hierarchical cluster analysis method is used.

## Spatially Constrained Clustering - SKATER approach

### Converting into SpatialPolygonsDataFrame

First, we need to convert `shan_sf` into SpatialPolygonsDataFrame. This is because SKATER function only support **sp** objects such as SpatialPolygonDataFrame.

The code chunk below uses [*as_Spatial()*](https://r-spatial.github.io/sf/reference/coerce-methods.html) of **sf** package to convert `shan_sf` into a SpatialPolygonDataFrame called `shan_sp`.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### Computing Neighbor List

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

Now, let's plot the neighbor network over the boundary map of Shan State.

```{r}
plot(shan_sp, 
     border=grey(.5))
plot(shan.nb, 
     coordinates(shan_sp), 
     col="blue", 
     add=TRUE)
```

### Computing minimum spanning tree

#### Calculating edge costs

Next, [*nbcosts()*](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** package is used to compute the cost of each edge. It is the distance between it nodes. This function compute this distance using a data.frame with observations vector in each node.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

We will incorporate these costs into a weights object in the same way as we did in the calculation of inverse of distance weights. In other words, we convert the neighbour list to a list weights object by specifying the just computed ***lcosts*** as the weights.

Note that we specify the *style* as **B** to make sure the cost values are not row-standardized.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### Computing minimum spanning tree

```{r}
shan.mst <- mstree(shan.w)
```

Let's check `shan.mst` class and dimensions.

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

```{r}
head(shan.mst)
```

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbor list is simplified to just one edge connecting each of the nodes, while passing through all the nodes.

```{r}
plot(shan_sp, border=gray(.5))
plot.mst(shan.mst, 
         coordinates(shan_sp), 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### Computing spatially constrained clusters using SKATER method

```{r}
clust6 <- skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

Let's examine the content of `clust6`.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary). This is followed by a detailed summary for each of the clusters in the edges.groups list. Sum of squares measures are given as `ssto` for the `total` and `ssw` to show the effect of each of the cuts on the overall criterion.

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. Parenthetially, we can also find this as the dimension of each vector in the lists contained in edges.groups. For example, the first list has node with dimension 12, which is also the number of observations in the first cluster.

```{r}
table(ccs6)
```

We can also plot the pruned tree that shows the five clusters on top of the townshop area.

```{r}
plot(shan_sp, border=gray(.5))
plot(clust6, 
     coordinates(shan_sp), 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005,
     add = TRUE)
```

### Visualising the clusters in choropleth map

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

Let's place the hierarchical clustering map and the spatially constrained hierarchical clustering map side by side to compare.

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5)

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=1, ncol=2)
```
